Plastic in the Urban Freshwater Environment

### 2021-11-12 code/analysis_code/exploration.Rmd
edit manuscript, expand modeling
removed EDA plots from manuscript, added RF and tree models, created grid for results

# load packages and data
```{r}
library(dplyr) #for data processing
library(here) #to set paths
library(tidyverse)
library(ggplot2)
library(GGally) #for producing correlation matrices

@@

@@

@@

mpdata$date <- factor(mpdata$date , levels=c("nov_20", "feb_21", "apr_21", "july_21"))

p1 <- mpdata %>% ggplot(aes(x = date, y = particles_l)) + 
  geom_boxplot(aes()) + 
  labs(title = "Microplastic Concentration by Sample Date", x = "Date", y = "Particles/Liter") + 
  scale_x_discrete(labels=c("Fall 2020",
                            "Winter 2021",
                            "Spring 2021",
                            "Summer 2021"))

@@

@@

I'm not seeing any obvious relationships here. Also, population values are appearing discrete rather than continous since there are a limited number of zip codes in the study area, and population was based on zip code.

@@

@@

Again, no obvious relationships here. How about if we looked with a logarithmic scale?

```{r}
p5 <- mpdata %>% ggplot(aes(x = e.coli.cfu, y = particles_l)) + 
  geom_point() + 
  theme(axis.text.x = element_text(angle = 90, hjust=1)) + 
  scale_x_continuous(trans='log10') +
  scale_y_continuous(trans='log10') + 
  labs(title = "Microplastic Concentration vs. E. coli CFU", x = "Log(CFU)", y = "Log(Particles/Liter)")

@@

Let's look at overall distribution of MP levels.

@@

@@

We see that most levels fall below 100 particles per liter. 
# Summary statistics
```{r}
summary(mpdata$particles_l)
```

These summary statistics give us a little more descriptive information about the dataset. The other variables that I wanted to look at particles per liter alongside were land cover. The process for looking at land cover will be a little more complicated - there a key for variable names corresponding to land types that I need to look closer at, and I am also not sure if the scale of the land cover data will be appropriate. Currently, it is at the census FIPS code level (approximately county level). Since most of these samples occur within that same area, we don't have other reference areas for comparison. I'll have to think about what I can do with this information.

@@

@@

@@

@@

plot(p10)
#save figure
figure_file = here("results","conc_bywatershed.png")
ggsave(filename = figure_file, plot=p10)
```
# Boxplot for individual sites throughout season
```{r}
p11 <- mpdata %>% ggplot(aes(x = site, y = particles_l, color = )) + 
  geom_boxplot()  +
  theme(axis.text.x = element_text(angle = 45, hjust=1)) +
  labs(title = "Microplastic Concentration by Site", x = "Site", y = "Particles/Liter") + ylim(0,500)

plot(p11)
#save figure
figure_file = here("results","conc_bysite.png")
ggsave(filename = figure_file, plot=p11)
```
# concentration over time by watershed
```{r}
# calculate mean concentration in each watershed for each sample date
mean_data <- group_by(mpdata, watershed, date) %>%
             summarise(particles_l = mean(particles_l, na.rm = TRUE))

@@

>>>>

# load packages and data
```{r}
library(dplyr) #for data processing
library(here) #to set paths
library(tidyverse)
library(ggplot2)
library(GGally) #for producing correlation matrices
library(gridExtra) #

@@

Let's start with some summary statistics. We want to see the overall summary of microplastic concentration - mean, median, range, etc.

@@

We have a total of 109 sites. Particles/L ranges from 16 to 1193, with a mean of 104 and a median of 67. This tells us the data is right-skewed (we have outliers on the upper end of the data).

Our observations can be categorized by sample date, since there were four different sample events that occurred. Let's see if there is any temporal relationship in the microplastic concentrations.

@@

mpdata$date <- factor(mpdata$date , levels=c("nov_20", "feb_21", "apr_21", "july_21")) #reorder dates to appear chronologically

# make boxplot for microplastic concentration by season
p1 <- mpdata %>% ggplot(aes(x = date, y = particles_l)) + 
  geom_boxplot(aes()) + 
  labs(title = "Microplastic Concentration by Sample Date", x = "Date", y = "Particles/Liter") + 
  scale_x_discrete(labels=c("Fall 2020",
                            "Winter 2021",
                            "Spring 2021",
                            "Summer 2021"))

@@

One of our main hypothesized predictors is population. Population size is a suppossed proxy for anthropogenic impact. We want to examine if microplastic levels increase in areas (census tracts) with greater populations.

@@

I'm not seeing any obvious relationships here. Also, population values are appearing discrete/in clusters rather than continous since there are a limited number of census tracts codes in the study area, and population was based on census tract.

Another way to visualize microplastic concentration vs population may be to log transform the microplastic levels. This would normalize the data, and the log transformation method has been used in previously published studies that analyze microplastic levels.

@@

This gives us a plot that is a little nicer to look at, but we still don't see a clear relationship between micrplastic level and local population.

Another potential predictor is e.coli colony forming units. Similar to population count, e.coli levels may be a proxy for human impact/nearby pollution sources. We want to see if there is a relationship between microplastic pollution and bacteria levels at the sample sites.

@@

Again, no obvious relationships here. We also see some pretty low values in this plot. How about if we looked with a logarithmic scale? Log transformation is often used for bacterial data, and can also be used with microplastic levels.

# log(MP) vs log(CFU) log transformed
```{r}
p5 <- mpdata %>% ggplot(aes(x = e.coli.cfu, y = particles_l)) + 
  geom_point() + 
  theme(axis.text.x = element_text(angle = 90, hjust=1)) + 
  scale_x_continuous(trans='log10') +
  scale_y_continuous(trans='log10') + 
  labs(title = "Microplastic Concentration vs. E. coli CFU", x = "Log(CFU)", y = "Log(Particles/Liter)")

@@

The log transformation may make these values easier to work with. Still no clear relationships, however.

Now, let's look at overall distribution of MP levels.

@@

Most observations fall below 300 particles/L, with a few large outliers.

@@

In this graph, we can more clearly see that most levels fall below 100 particles per liter.

Another way to look at relationships among predictors and outcome is through correlation matrices. These allow us to view multiple relationships at once.

@@

There are no bright red or bright blue boxes here, so we don't neccesarily have strong relationships among variables. We see that temperature and e.coli are positively correlated, as are temperature and turbidity. There is a positive relationship between turbidity and population. The correlation between particles/L and population is very weak - nearly completely neutral. The only moderate correlation is between turbidity and particles/L. There appears to be a very weak negative correlation between e.coli and particles/L which is unexpected. However, we are basing this off of colors and not specific coefficients or p-values, so this is a vague estimate.

Since we have a lot of different variables for land use, the correlation matrix is a good option to explore these and view many at the same time.

@@

In this plot, we are mainly looking at the row containing particles/L (the very bottom row). We do not see any strong correlations between microplastic level and local land use.

We also may want to see the variability of microplastic concentration at each individual site (rather than aggregate concentration) througout the seasons/sample dates. Let's take a look.

@@

In this plot, we see that some sites experienced small differences in concentration between seasons, while other experienced large differences between seasons.

Similarly, let's see if there are any distictions between watersheds, since we can group the sample sites based on what watershed they are located in.

@@

plot(p10)
#save figure
figure_file = here("results","conc_bywatershed.png")
ggsave(filename = figure_file, plot=p10)
```

A few watersheds exhibit more variation/range in microplastic level than other watersheds. This also may be due to the number of sites located in each watershed (some watersheds contain very few sites while others contain many). It may be useful to include an "n =" on this graph in the future.

Instead of viewing a dot/point for each sample data, let's view the concentrations at each site throughout the year as a boxplot to visualize variability, averages, etc.

# Boxplot for individual sites throughout season
```{r}
p11 <- mpdata %>% ggplot(aes(x = site, y = particles_l, color = )) + 
  geom_boxplot()  +
  theme(axis.text.x = element_text(angle = 45, hjust=1)) +
  labs(title = "Microplastic Concentration by Site", x = "Site", y = "Particles/Liter") + ylim(0,500)

plot(p11)
#save figure
figure_file = here("results","conc_bysite.png")
ggsave(filename = figure_file, plot=p11)
```

There are a few sites that have quite a large range of values, though the majority of the sites have a small to moderate range. It must also be noted that not all sample sites were accounted for at each sample date. So, some sites may only have 1 or 2 observations, while other sites may have 4 observations.

I want to visualize the change (or lack of change) in MP concentration over time. However, that is a lot to take in looking at individual sites. Let's view the average concentrations in each watershed for each sample date. Then, we can look at those watershed averages over time throughout the study period.

# concentration over time by watershed
```{r}
# calculate mean concentration in each watershed for each sample date
mean_data <- group_by(mpdata, watershed, date) %>%
             summarise(particles_l = mean(particles_l, na.rm = TRUE))

@@

Looking at the data this way, we do see somewhat of a pattern. It appears that the range of microplastic concentration decreases from Fall 2020 to Summer 2021. This may be due to some particularly large outliers in the earlier sample dates. Also, we may consider human error. Since I counted these microplastics myself, it may be that I initially over-counted, including false positives (particles that I mistook for microplastics, when they may have been organic matter, etc). Essentially, methods may have improved/refined over time.

### 2021-11-12 code/analysis_code/modeling.Rmd
edit manuscript, expand modeling
removed EDA plots from manuscript, added RF and tree models, created grid for results

In this portion of the analysis, we will use the LASSO method for regularization of the model.

# load packages and data
```{r}
library(ggplot2) #for plotting
library(broom) #for cleaning up output from lm()
library(here) #for data loading/saving
library(tidymodels) #for modeling
library(rpart)
library(glmnet)
library(ranger)
library(rpart.plot)  # for visualizing a decision tree
library(vip)         # for variable importance plots

@@

#model predictions from tuned model vs actual outcomes
lasso_pred_plot <- ggplot(lasso_resid, aes(x = particles_l, y = .pred)) + geom_point() + 
  labs(title = "Predictions vs Actual Outcomes: LASSO", x = "Particles/L Outcome", y = "Particles/L Prediction")
lasso_pred_plot

#plot residuals vs predictions
lasso_resid_plot <- ggplot(lasso_resid, aes(y = .resid, x = .pred)) + geom_point() + 
  labs(title = "Predictions vs Residuals: LASSO", x = "Particles/L Prediction", y = "Residuals")
lasso_resid_plot #view plot

@@

#plot for how the number of predictors included in the LASSO model changes with the tuning parameter
```{r}
x <- lasso_fit$fit$fit$fit 
plot(x, "lambda")

@@

@@

>>>>

In this portion of the analysis, we will use the LASSO method for regularization of the model. We'll also try out a decision tree

# load packages and data
```{r}
library(ggplot2) #for plotting
library(broom) #for cleaning up output from lm()
library(here) #for data loading/saving
library(tidymodels) #for modeling
library(rpart)
library(glmnet)
library(ranger)
library(rpart.plot)  # for visualizing a decision tree
library(vip)         # for variable importance plots
library(gridExtra)

@@

#model predictions from tuned model vs actual outcomes
lasso_pred_plot <- ggplot(lasso_resid, aes(x = particles_l, y = .pred)) + geom_point() + 
  labs(title = "LASSO: Pred vs Outcomes", x = "Particles/L Outcome", y = "Particles/L Prediction")
lasso_pred_plot

#plot residuals vs predictions
lasso_resid_plot <- ggplot(lasso_resid, aes(y = .resid, x = .pred)) + geom_point() + 
  labs(title = "LASSO: Pred vs Residuals", x = "Particles/L Prediction", y = "Residuals")
lasso_resid_plot #view plot

@@

# plot for how the number of predictors included in the LASSO model changes with the tuning parameter
```{r}
x <- lasso_fit$fit$fit$fit 
plot(x, "lambda")

@@

# Decision tree
Next, we'll try out a decision tree model

## Decision tree
```{r}
#going based off of tidymodels tutorial: tune parameters
#since we already split our data into test and train sets, we'll continue to use those here. they are `train_data` and `test_data`

#model specification
tune_spec <- 
  decision_tree(
    cost_complexity = tune(),
    tree_depth = tune()
  ) %>% 
  set_engine("rpart") %>% 
  set_mode("regression")
tune_spec

#tuning grid specification
tree_grid <- grid_regular(cost_complexity(),
                          tree_depth(),
                          levels = 5)
tree_grid

#cross validation
set.seed(123)
cell_folds <- vfold_cv(train_data)

#workflow
set.seed(123)

tree_wf <- workflow() %>%
  add_model(tune_spec) %>%
  add_recipe(MP_rec)

#model tuning with `tune_grid()`
tree_res <- 
  tree_wf %>% 
  tune_grid(
    resamples = cell_folds,
    grid = tree_grid
    )
tree_res %>% collect_metrics()
#Here we see 25 candidate models, and the RMSE and Rsq for each
tree_res %>% autoplot() #view plot

#select the best decision tree model
best_tree <- tree_res %>% select_best("rmse")
best_tree #view model details

#finalize model workflow with best model
tree_final_wf <- tree_wf %>%
  finalize_workflow(best_tree)

#fit model
tree_fit <- 
  tree_final_wf %>% fit(train_data)
tree_fit

@@

### Decision tree plots
```{r}
#diagnostics
autoplot(tree_res)
#calculate residuals - originally got stuck trying out lots of different methods for this. took inspiration from Zane's code to manually calculate residuals rather than using some of the built in functions that I could not get to cooperate
tree_resid <- tree_fit %>%
  augment(train_data) %>% #this will add predictions to our df
  select(.pred, particles_l) %>%
  mutate(.resid = particles_l - .pred) #manually calculate residuals

#model predictions from tuned model vs actual outcomes
tree_pred_plot <- ggplot(tree_resid, aes(x = particles_l, y = .pred)) + geom_point() + 
  labs(title = "Decision Tree: Pred vs Outcomes", x = "Particles/L Outcome", y = "Particles/L Prediction")
tree_pred_plot

#plot residuals vs predictions
tree_resid_plot <- ggplot(tree_resid, aes(y = .resid, x = .pred)) + geom_point() + 
  labs(title = "Decision Tree: Pred vs Residuals", x = "Particles/L Prediction", y = "Residuals")
tree_resid_plot #view plot

#compare to null model
metrics_null_train #view null RMSE for train data
tree_res %>% show_best(n=1) #view RMSE for best decision tree model
```

The best decision tree has an RMSE of 102.2, which is better than the null model RMSE of 109. However, the decision tree did not perform as well well as the Lasso regularized model, which had an RMSE of about 100.

# Random forest
Finally, let's try out a random forest.

### split data into train and test subsets with only COMPLETE CASES
```{r}
data %>% is.na() %>% summary() #check out missing cases
complete <- data %>% na.omit() #save complete cases
complete <- complete %>% select(-"visual_score")

# set seed for reproducible analysis (instead of random subset each time)
set.seed(123)
#subset 3/4 of data as training set
data_split <- initial_split(complete, 
                            prop = 7/10, 
                            strata = particles_l) #stratify by MP concentration for balanced outcome

#save sets as data frames
train_data <- training(data_split)
test_data <- testing(data_split)

#reset recipe with adjusted, complete data
MP_rec <- recipe(particles_l ~ ., data = train_data) %>% step_dummy(all_nominal())
#reset cell_folds
set.seed(123)
cell_folds <- vfold_cv(train_data)
```

## Random forest
```{r}
#based on tidymodels tutorial: case study
library(parallel)
cores <- parallel::detectCores()
cores
#model specification
r_forest <- rand_forest(mtry = tune(), min_n = tune(), trees = tune()) %>% set_engine("ranger", num.threads = cores) %>% set_mode("regression")

#set workflow
r_forest_wf <- workflow() %>% add_model(r_forest) %>% add_recipe(MP_rec)

#tuning grid specification
rf_grid  <- expand.grid(mtry = c(3, 4, 5, 6), min_n = c(40,50,60), trees = c(500,1000)  )
#what we will tune:
r_forest %>% parameters()

#tuning with CV and tune_grid

r_forest_res <- 
  r_forest_wf %>%
  tune_grid(resamples = cell_folds, 
            grid = rf_grid, 
            control = control_grid(save_pred = TRUE), 
            metrics = metric_set(rmse))

#view top models
r_forest_res %>% show_best(metric = "rmse")

#view plot of models performance
autoplot(r_forest_res)

#select best model
rf_best <- r_forest_res %>% select_best(metric = "rmse")
rf_best

#finalize workflow with top model
rf_final_wf <- r_forest_wf %>% finalize_workflow(rf_best)

#fit model with finalized WF
rf_fit <- rf_final_wf %>% fit(train_data)
```
Here, we get an error message that all models failed, note: missing value in visual_score. Let's go back and use only complete cases for the random forest.

The issue was in "cell_folds" which had originated earlier in the Rmd and didn't include the updated complete case dataset. Added a line of code to reset cell_folds at this stage of the Rmd in order to use complete cases only. We still get some error messages but are able to run the model.

The best random forest models have RMSEs of about 108. This is slightly better than the null model, but not significantly. The random forest model does not perform as well as the decision tree or LASSO model.

### Random forest plots
```{r}
#diagnostics
autoplot(r_forest_res)
#calculate residuals
rf_resid <- rf_fit %>%
  augment(train_data) %>% #this will add predictions to our df
  select(.pred, particles_l) %>%
  mutate(.resid = particles_l - .pred) #manually calculate residuals

#model predictions from tuned model vs actual outcomes
rf_pred_plot <- ggplot(rf_resid, aes(x = particles_l, y = .pred)) + geom_point() + 
  labs(title = "Random Forest: Pred vs Outcomes", x = "Particles/L Outcome", y = "Particles/L Prediction")
rf_pred_plot

#plot residuals vs predictions
rf_resid_plot <- ggplot(rf_resid, aes(y = .resid, x = .pred)) + geom_point() + 
  labs(title = "Random Forest: Pred vs Residuals", x = "Particles/L Prediction", y = "Residuals")
rf_resid_plot #view plot

#compare to null model
metrics_null_train #view null RMSE for train data
r_forest_res %>% show_best(n=1) #view RMSE for best decision tree model
```

```{r}
model_grid <- grid.arrange(lasso_pred_plot, lasso_resid_plot, rf_pred_plot, rf_resid_plot, tree_pred_plot, tree_resid_plot, ncol = 3)

ggsave(model_grid, filename = here("results", "model_grid.png"))
```

I'll need to go back and edit this image later so that the titles aren't cut off and the axis scales match up properly so the plots look right. I might also want to add a 45 degree diagonal line to the outcomes vs predictions plot, and some other features to improve appearance.

### 2021-11-12 products/manuscript/Manuscript.Rmd
edit manuscript, expand modeling
removed EDA plots from manuscript, added RF and tree models, created grid for results

There are similar microplastic levels throughout the watersheds within the Upper Oconee. Some watersheds experienced greater variation in microplastic levels than other watersheds. Figure \@ref(fig:conc-bywatershed) shows the microplastic concentrations by watershed.

```{r conc-bywatershed,  fig.cap='Watershed Microplastic Concentrations', echo=FALSE}
knitr::include_graphics("../../results/conc_bywatershed.png")
```

@@

### Predictors
Population, land cover/use, and bacteria levels are hypothesized predictors of microplastic concentration. Figure \@ref(fig:particlesvpop) and Figure \@ref(fig:logmpvcfu) demonstrate the relationship between microplastic concentration and population and microplastic concentration and bacteria levels (CFU/L), respectively.

```{r particlesvpop,  fig.cap="Particles/L vs Population", echo=FALSE}
knitr::include_graphics("../../results/particlesvpop.png")
```

@@

Figure \@ref(fig:matrix) and Figure \@ref(fig:landmatrix) show correlation matrices for the hypothesized predictor and for the different categories of land use.

```{r matrix,  fig.cap='Predictor matrix', echo=FALSE}
knitr::include_graphics("../../results/matrix.png")
```

```{r landmatrix,  fig.cap='Land cover matrix', echo=FALSE}
knitr::include_graphics("../../results/landmatrix.png")
```

@@

Preliminary modeling reveals that there is not a strong relationship between microplastic concentration and population level. Figure \@ref(fig:m-concvpop-no-outlier) demonstrates a linear model fit.

```{r m-concvpop-no-outlier,  fig.cap='Concentration vs Population Linear Model', echo=FALSE}
knitr::include_graphics("../../results/m_concvpop_no_outlier.png")
```

@@

>>>>

@@

### Predictors
Population, land cover/use, and bacteria levels are hypothesized predictors of microplastic concentration. Figure \@ref(fig:logparticlesvpop) and Figure \@ref(fig:logmpvcfu) demonstrate the relationship between microplastic concentration and population and microplastic concentration and bacteria levels (CFU/L), respectively.

```{r logparticlesvpop,  fig.cap="Particles/L vs Population", echo=FALSE}
knitr::include_graphics("../../results/logparticlesvpop.png")
```

@@

@@

Preliminary modeling reveals that there is not a strong relationship between microplastic concentration and population level. Figure \@ref(fig:m-logconcvpop) demonstrates a linear model fit.

```{r m-logconcvpop-no-outlier,  fig.cap='Concentration vs Population Linear Model', echo=FALSE}
knitr::include_graphics("../../results/m_logconcvpop.png")
```

@@

Beyond the basic linear model, we have applied additional methods to improve model performance, including LASSO regularization and building decision trees and random forests for model comparison. The predictions, outcomes, and residuals resulting from each type of plot are demonstrated in Figure \@ref(fig:model-grid)

```{r model-grid, fig.cap='Model Quality', echo=FALSE}
knitr::include_graphics("../../results/model_grid.png")
```
